---
title: "Tidyverse for COVID19 Data Analysis"
author: "J. Wall"
date: "`r Sys.Date()`"
output:
  word_document: default
  pdf_document: default
  html_document:
    df_print: paged
  powerpoint_presentation:
    slide_level: 3 # use this to override default
    reference_doc: my_template.pptx
---

```{r setup, include=FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(lubridate)
library(broom)
```



# Get and clean worldwide Data

### Data from Johns Hopkins github
```{r filenames, message = FALSE}
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c("time_series_covid19_confirmed_global.csv",
                "time_series_covid19_deaths_global.csv",
                "time_series_covid19_confirmed_US.csv",
                "time_series_covid19_deaths_US.csv")
urls <- str_c(url_in,file_names)
```

### Check urls
```{r JHU_urls}
urls
```

### Tidy global data

```{r tidy_global}
global_confirmed <- read_csv(urls[1]) %>%
    pivot_longer(cols = -c(`Province/State`, `Country/Region`, Lat, Long), names_to = "Date", 
               values_to = "Confirmed_cases")
global_deaths <- read_csv(urls[2]) %>%
    pivot_longer(cols = -c(`Province/State`, `Country/Region`, Lat, Long), names_to = "Date", 
               values_to = "Deaths")
    
global <- global_confirmed %>% full_join(global_deaths) %>%
  rename(Country_Region = `Country/Region`, Province_State = `Province/State`) %>%
  mutate(Date = mdy(Date))
```

### fix problems
```{r fix_global, results = "asis"}
knitr::kable(summary(global))
# we find there are entries of -1 for Diamond Princess
global <- global %>% filter(Confirmed_cases >= 0, Deaths >= 0)
```

### Join population data to the dataset

```{r get_pop}
uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
uid <- read_csv(uid_lookup_url) %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))
global <- global %>% 
  left_join(uid, by = c("Province_State", "Country_Region")) %>%
  select(-c(UID, FIPS)) %>%
  select(Province_State, Country_Region, Date,
         Confirmed_cases, Deaths, Population,
         Lat, Long)
```

### Get and tidy US data
```{r tidy_US}
US_confirmed <- read_csv(urls[3]) %>%
    pivot_longer(cols = -(UID:Combined_Key), names_to = "Date", values_to = "Confirmed_cases") %>%
    select(Admin2:Confirmed_cases) %>%
    mutate(Date = mdy(Date))
US_deaths <- read_csv(urls[4]) %>%
    pivot_longer(cols = -(UID:Population), names_to = "Date", values_to ="Deaths") %>%
    select(Admin2:Deaths) %>%
    mutate(Date = mdy(Date))
```

### Join deaths and cases
```{r join_US_deaths_cases}
US <-  US_deaths %>%
  full_join(US_confirmed, 
            by = c("Combined_Key", "Date", 
                   "Admin2", "Province_State", 
                   "Country_Region")) %>%
    rename(Long = Long_.x, Lat = Lat.x)  %>%
    select(Admin2, Province_State, Country_Region, 
           Lat, Long, Population, Date, Confirmed_cases, Deaths)
```

### US data so far
```{r us_data_intermediate}
US %>% filter(Province_State == "New York") %>%
  select(Admin2, Province_State, Confirmed_cases, Deaths) %>%
  head(n = 4)
#Note that what we now have is county level data within each state. It would be nice to have data totaled for each state.
```

## DMV data

### get DMV data
```{r dmv_data}
dmv_data <-
  US %>%
  mutate(state = factor(Province_State), 
         county = factor(Admin2)) %>%
  filter(state %in% c("Maryland", "Virginia", "District of Columbia")) %>%
  filter(county %in% 
           c("Anne Arundel", "Montgomery", "Howard", "Frederick", "Prince George's",
             "Charles", "District of Columbia", "Alexandria", "Arlington", "Fairfax",
             "Loudoun", "Prince William")) %>%
  select(-c(Admin2, Province_State, Country_Region)) %>%
  mutate(Deaths_per_mill = 1000000 * Deaths / Population) %>%
  select(state, county, Date, Confirmed_cases, Deaths,
         Deaths_per_mill, Population, Lat, Long) 
```

### Plot DMV cases by county
```{r dmv_over_time, fig.width = 10}
dmv_data %>%
  filter(Confirmed_cases > 0) %>%
  ggplot(aes(x = Date, y = Confirmed_cases, group = county, col = county)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ state) +
  scale_y_log10() +
  scale_color_brewer(palette = "Paired") + 
  guides(color = guide_legend(ncol = 4)) +
    theme(legend.position="bottom", axis.text.x = element_text(angle = 90))
```

### Compute cases & deaths in DMV 
```{r exercise3_solution}
DMV_totals <- dmv_data %>% 
  group_by(Date) %>%
  summarize(total_cases = sum(Confirmed_cases), 
            total_deaths = sum(Deaths))

paste0("total deaths = ", as.character(max(DMV_totals$total_deaths)) )

paste0("total cases = ", as.character(max(DMV_totals$total_cases) ))
# Latest data from:
paste0("latest data from: ", as.character( max(DMV_totals$Date) ))
```

### Visualize DMV cases and deaths
```{r plot_DMV_deaths}
DMV_totals %>%
  filter(total_cases > 0) %>%
  ggplot(aes(x = Date, y = total_cases)) +
  geom_line() +
  geom_point() +
  labs(title = "Confirmed Cases of COVID19 in DMV", x = NULL, y = NULL)

DMV_totals %>%
  filter(total_deaths > 0) %>%
  ggplot(aes(x = Date, y = total_deaths)) +
  geom_line(color = "red") +
  geom_point() +
  labs(title = "Deaths due to COVID19 in DMV", x = NULL, y = NULL)
```

### both on one graph 
```{r DMV_plot_both}
DMV_totals %>%
  filter(total_cases > 0) %>%
  ggplot(aes(x = Date, y = total_cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = total_deaths, color = "deaths"),
            data = DMV_totals %>% filter(total_deaths > 0)) +
  geom_point(aes(y = total_deaths, color = "deaths"),
             data = DMV_totals %>% filter(total_deaths > 0)) +
  scale_y_log10() + 
  labs(title = "COVID19 in DMV", x = NULL, y = NULL)
```

## State analysis

### Compute state totals

```{r state_total_data}
US_by_state <- US %>%
  group_by(Province_State, Country_Region, Date) %>%
  # add up counties and population
  summarize(Confirmed_cases = sum(Confirmed_cases), 
            Deaths = sum(Deaths), Lat = median(Lat), 
            Long = median(Long), Population = sum(Population)) %>%
  select(Province_State, Country_Region, Date,
         Confirmed_cases, Deaths, Population,
         Lat, Long) %>%
  ungroup()
```

### State data now
```{r state_total_show}
US_by_state %>% head(n = 3) %>% 
  select(Province_State, Date, Confirmed_cases, 
         Deaths, Population, Country_Region)
```
### Order deaths and cases by state

```{r deaths_per_state}
US_state_totals <- US_by_state %>% 
  group_by(Province_State) %>% 
  summarize(cases = max(Confirmed_cases),
            deaths = max(Deaths),
            population = max(Population)) %>%
  filter(cases > 0) %>%
  mutate(deaths_per_mill = 1000000 * deaths / population) %>%
  arrange(desc(cases))
```

### View list by states

```{r view_state_totals, results = "asis"}
knitr::kable(US_state_totals %>% slice(1:10))
```

### Totals for US
```{r us_total_deaths}
# total deaths 
paste0("total US deaths = ", as.character(sum(US_state_totals$deaths)) )

# total cases
paste0("total US cases = ", as.character(sum(US_state_totals$cases)) )

# Latest data from:
paste0("latest data from: ", as.character(max(US_by_state$Date)) )
```


## Visualizing the state data

```{r usmap, warning = FALSE}
library(usmap)
US_data <- US_state_totals %>% 
  mutate(state = Province_State) %>% 
  filter(!is.na(deaths_per_mill))
```

### Plot states
```{r plot_state_colors, fig.width = 9}
plot_usmap(data = US_data, 
           values = "deaths_per_mill",
           color = "black") + 
    scale_fill_gradient(name = "Deaths per million",
                      low = "yellow", high = "red") +
  theme(legend.position = "right")
```

### partition deaths per million into 10 equal ranges.

```{r cut_deaths, warning = FALSE}
US_data <- US_data %>%
  filter(population > 0) %>%
  mutate(death_group = cut(deaths_per_mill,
                        breaks = seq(min(deaths_per_mill),
                                     max(deaths_per_mill),
                                     length.out = 10), 
                        include.lowest = TRUE,
                        right = FALSE, 
                        ordered_result = TRUE) )
```

### Plot 10 levels
```{r plot_10_levels, fig.width = 9}
plot_usmap(data = US_data, 
           values = "death_group",
           color = "black") + 
  scale_fill_discrete(name = "Deaths_per_million") +
            theme(legend.position = "right")
```

### Add states to global
```{r expand_US}
#Replace the US observations in the global dataset with the US data
exp_global <- global %>% 
  # remove the US total data from the dataset
  filter(Country_Region != "US") %>% 
  # add on the totals by state
  bind_rows(US_by_state)
```

### Add continents
```{r add_continents}
library(countrycode)
temp <- countrycode(exp_global$Country_Region,
                    origin = "country.name",
                    destination = "continent")
Confirmed <- exp_global %>%
  mutate(continent = temp) %>%
  mutate(continent = case_when(
    Country_Region == "Cruise Ship"~"Cruiseship",
    Country_Region == "Diamond Princess"~"Cruiseship",
    Country_Region == "MS Zaandam"~"Cruiseship",
    Country_Region == "Kosovo" ~ "Europe",
    TRUE ~ continent)) %>%
  # create a Country_State combining Province_State & Country_Region
  unite(Country_State, c(Country_Region, Province_State), 
        na.rm = TRUE, remove = FALSE)
Confirmed %>% filter(is.na(continent))
```
### Compute Deaths per million population 
```{r deaths_per_million}
Confirmed <- Confirmed %>%
  mutate(Deaths_per_mill = 1000000 * Deaths / Population) %>%
  select(Country_State, Date, Confirmed_cases, 
         Deaths, Deaths_per_mill, continent, 
         Population, Lat, Long, everything()) %>%
  filter(Confirmed_cases > 0) # leave off rows w/o cases
```

### Country/State's w/ most cases 
```{r country_state_totals}
Confirmed_totals <- Confirmed %>% 
  group_by(Country_State, Province_State, Country_Region, continent) %>%
  summarize(Confirmed_cases = max(Confirmed_cases), 
            Deaths = max(Deaths),
            Deaths_per_mill = max(Deaths_per_mill),
            Date_first_case = min(Date),
            Population = max(Population)) %>%
  ungroup() %>%
  select(Country_State, Date_first_case, Deaths_per_mill,
         Deaths, Confirmed_cases, Population, everything())
```

### Worldwide totals to date

```{r world_total_deaths}
# total deaths 
paste0("total worldwide deaths = ", as.character(sum(Confirmed_totals$Deaths)) )

# total cases
paste0("total worldwide cases = ", as.character(sum(Confirmed_totals$Confirmed_cases)) )

# average deaths per million to date
df <- Confirmed_totals %>%
  summarize(death_rate = 1000000 * sum(Deaths, na.rm = TRUE) / sum(Population, na.rm = TRUE))

paste0("total worldwide deaths per million to date = ", as.character(sum(df$death_rate)) )

# Latest data from:
paste0("latest data from: ", as.character(max(US_by_state$Date)) )
```

### Top 25 and top 100
```{r top_25_50, results = "asis"}
Top_25 <- Confirmed_totals %>% 
  arrange(desc(Confirmed_cases)) %>%
  slice(1:25) %>%
# Top_25 %>% 
  select(Country_State, continent,
          Confirmed_cases, Deaths,Deaths_per_mill) 
knitr::kable(Top_25 %>% slice(1:8))
Top_100 <- Confirmed_totals %>% 
  arrange(desc(Confirmed_cases)) %>%
  slice(1:100)
```

### Get data for top 25

```{r plot_top_25}
# grab top 25 country / states for graphing
Top_25_states <- Top_25$Country_State
Top_25_data <- Confirmed %>% 
  filter(Country_State %in% Top_25_states) %>%
  select(Country_State, continent, Date, Confirmed_cases, 
         Deaths, Deaths_per_mill)
```

### Graph top 25 
```{r graph_top25, fig.width=10, warning=FALSE}
Top_25_data %>% filter(Confirmed_cases > 0) %>%
  ggplot(aes(x = Date, y = Confirmed_cases, 
             group = Country_State, 
             color = Country_State)) +
    geom_line() +
    facet_wrap(~continent, scales = "free") +
    scale_y_log10() +
    labs(title = "Confirmed Cases - top 25", x = NULL, y = NULL) +
    guides(color = guide_legend(ncol = 6)) +
    theme(legend.position="bottom",axis.text.x = element_text(angle = 90))
Top_25_data %>% filter(Deaths > 0) %>%
  ggplot(aes(x = Date, y = Deaths, 
             group = Country_State, 
             color = Country_State)) +
    geom_line() +
    facet_wrap(~continent, scales = "free") +
    scale_y_log10() +
    labs(title = "Deaths - top 25", x = NULL, y = NULL) + 
    guides(color = guide_legend(ncol = 6)) +
    theme(legend.position="bottom",axis.text.x = element_text(angle = 90))
Top_25_data %>% filter(Deaths > 0) %>%
  ggplot(aes(x = Date, y = Deaths_per_mill, 
             group = Country_State, 
             color = Country_State)) +
    geom_line() +
    labs(title = "Deaths per million population", x = NULL,
         y = NULL) +
    facet_wrap(~continent, scales = "free") +
    scale_y_log10() +
    guides(color = guide_legend(ncol = 6)) +
    theme(legend.position="bottom",axis.text.x = element_text(angle = 90))
```


## Scandinavia analysis

```{r scandinavia, results = "asis"}
# look at cases in Scandinavia since Sweden has not shut down their economy like other countries have.  What impact has this had on death rates?
Scandinavia <- Confirmed %>% 
  filter(Country_State %in% c("Sweden", "Denmark", "Finland")) %>%
  select(Country_State, Date, Confirmed_cases, Deaths, Deaths_per_mill, everything()) %>%
  mutate(Country_State = factor(Country_State)) 

Scandinavia %>% 
  filter(Deaths > 0) %>%
    ggplot(aes(x = Date, y = Deaths, color = Country_State)) +
  geom_point() + geom_line() +
  labs(title = "Deaths", x = NULL, y = NULL) +
  scale_y_log10()
Scandinavia %>% 
  filter(Deaths > 0) %>%
    ggplot(aes(x = Date, y = Deaths_per_mill, color = Country_State)) +
  geom_point() + geom_line() +
  labs(title = "Deaths per million", x = NULL, y = NULL) +
  scale_y_log10()  
Scand_summ <- Scandinavia %>% group_by(Country_State) %>%
  summarize(Max_Deaths_per_million =
              max(Deaths_per_mill),
            Total_cases = max(Confirmed_cases),
            Total_deaths = max(Deaths),
            Population = max(Population))
knitr::kable(Scand_summ)
```


# Modeling

### Hubei province of China and the Diamond Princess cruise ship.

```{r slowed}
Slowed_cases <- Confirmed %>% 
  filter(Country_State %in% c("China_Hubei", "Diamond Princess")) 
Slowed_cases %>%
  ggplot(aes(x = Date, y = Confirmed_cases)) +
    geom_point() +
  facet_wrap(~ Country_State, scales = "free") 
```

### Fitting a sigmoid function.
```{r nls}
# thanks to http://kyrcha.info/2012/07/08/tutorials-fitting-a-sigmoid-function-in-r
# function needed for visualization purposes
sigmoid = function(x, params) {
  params[1] / (1 + exp(-params[2] * (x - params[3])))
}

x = 1:53
y = c(0,0,0,0,0,0,0,0,0,0,0,0,0,0.1,0.18,0.18,0.18,0.33,0.33,0.33,0.33,0.41,0.41,0.41,0.41,0.41,0.41,0.5,0.5,0.5,0.5,0.68,0.58,0.58,0.68,0.83,0.83,0.83,0.74,0.74,0.74,0.83,0.83,0.9,0.9,0.9,1,1,1,1,1,1,1)
df <- tibble(x = x, y = y)    
# fitting code
fitmodel <- nls(y ~ a /(1 + exp(-b * (x - c))), data = df,
                start = list(a = 1, b = 0.5, c = 25))
```

### Plot model and data for sigmoid example
```{r nls2}
# visualization code
# get the coefficients using the coef function
params=coef(fitmodel)
    
df$y2 <- sigmoid(x, params)
df %>% ggplot(aes(x, y2)) + geom_line() + geom_point(y = y)
```

### More generalized sigmoid function 

$$y(t) = \frac{K}{\left(1 + Q*e^\left(-B*\left(t - t0\right)\right)\right)^\frac{1}{v}}$$ 

Goal: use that form for Hubei and fit a sigmoid function to that data. 

### Start simply

$$y(t) =  \frac{K}{\left(1 + e^\left(-B*\left(t - t0\right)\right)\right)}$$

### Get, visualize, model Hubei data

```{r Hubei_simple_sigmoid}
library(broom)
Hubei_cases <- Confirmed %>% 
  filter(Country_State == "China_Hubei") %>%
  mutate(date_int = unclass(Date))
Hubei_cases %>%
  ggplot(aes(x = Date, y = Confirmed_cases)) +
    geom_point()  
# first with simplified sigmoid function
sigmoid <- function(x, params) {
  params[1] / (1 + exp(-params[2] * (x - params[3])))
}
mod1 <- nls(Confirmed_cases ~ K /(1 + exp(-B * (date_int - t0))), 
         data = Hubei_cases,
         start = list(K = 60000, B = 0.5, t0 = 18300))
params <- coef(mod1)
```

### Plot Hubei model and data
```{r Hubei_plot}
Hubei_cases <- Hubei_cases %>% 
  mutate(pred = sigmoid(date_int, params)  ) %>%
  select(Confirmed_cases, pred, Date, date_int, everything())

Hubei_cases %>% 
  ggplot(aes(Date, pred)) + 
  geom_line() + 
  geom_point(aes(y = Confirmed_cases))
```

### Model summary
```{r mod1_stats}
summary(mod1)
glance(mod1)
```

### One more step

$$y(t) = \frac{K}{\left(1 + e^\left(-B*\left(t - t0\right)\right)\right)^\frac{1}{v}}$$ 

### More complex model on Hubei

```{r Hubei_sigmoid}
sigmoid_gen <- function(x, params) {
  params[1] /
    ( (1 +  exp(-params[2] * (x - params[3]))) ) ^ (1 / params[4])
}

startK <- max(Hubei_cases$Confirmed_cases)
mod2 <- nls(Confirmed_cases ~ K /( (1 +  exp(-B * (date_int - t0))) ) ^(1/v) , 
                data = Hubei_cases,
                start = list(K = startK,  B = .25, t0 = 18300, v = 1) ) 
```

### Visualize both Hubei models
```{r viz_models}
# get the coefficients using the coef function
params <- coef(mod2)
    
Hubei_cases <- Hubei_cases %>% 
  mutate(pred2 = sigmoid_gen(date_int, params) ) %>%
  select(Confirmed_cases, pred2, pred, Date, date_int, everything())

Hubei_cases %>% 
  ggplot(aes(x = Date)) + 
  geom_line(aes(y = pred2), color = "red") + 
  geom_line(aes(y = pred), color = "black") +
  geom_point(aes(y = Confirmed_cases))
# use broom::glance to look at model results
summary(mod2)
glance(mod2)
```
### Our plan of attack

* use a map function and nested grouped data to develop models out of this more complicated model family for the Diamond Princess and most of the China provinces
* Add the model to the nested data
* Use `broom::tidy()` to add the coefficients of the models to the dataset
* Use pivot_wider to make these coefficients into columns.

### Nest China and Princess data

```{r many_models, warning = FALSE}
by_country <- Confirmed %>%
  filter(Country_State != "China_Beijing",
         Country_State != "China_Shanxi", 
         Country_State != "China_Tibet") %>%
   filter(Country_Region == "China" | Country_State == "Diamond Princess") %>% 
  #filter(Confirmed_cases > 200) %>%
  mutate(date_int = unclass(Date)) %>%
  group_by(Country_State, Province_State, Country_Region, 
           continent, Lat, Long) %>%
  nest()
# by_country$data[[1]]
by_country
```

### Add models to nested data
```{r add_models, warning = FALSE}
country_mod2 <- function(df){
  startK <- max(df$Confirmed_cases)
  nls(Confirmed_cases ~ K /( (1 +  exp(-B * (date_int - t0))) ) ^(1/v) , 
        data = df,
        start = list(K = startK,  B = .25, t0 = 18300, v = 1),
        control = list(maxiter = 1000, warnOnly = TRUE))
}

by_country <- by_country %>%
  mutate(model = map(data, country_mod2),
         tm = map(model, broom::tidy)) %>%
  unnest(tm) %>%
  select(Country_State, Lat, Long, term, estimate,
         `std.error`, statistic, p.value, everything())
```

### Look at results

```{r int_model_results}
by_country
```

### Make data per province
```{r make_table}
province_data <- by_country %>%
  ungroup() %>%
  select(Province_State, Lat, Long, term, estimate, data) %>%
  pivot_wider(names_from = term, values_from = estimate) %>%
  mutate(t0_date = as_date(t0))
province_data %>% arrange(t0_date) %>% print(n = Inf)
```

### Join with province totals
```{r province_totals}
pd <- province_data %>% left_join(Confirmed_totals) %>%
  filter(!is.na(Province_State))
pd %>% print(n = Inf, width = Inf)
```
